---
title: "Assessment"
output: html_document
date: "2025-10-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### The impact of COVID-19 pandemic on Type 2 diabetes medication prescribing patterns, and how prescription patterns vary across deprived areas?

##### During the COVID-19 pandemic and its associated lockdowns, population health was significantly affected. Reduced physical activity, disruptions to daily routines, and increased comfort eating contributed to rising concerns about weight gain and metabolic health. Given these conditions, an increase in obesity and consequently Type 2 diabetes might be expected. This project examines how Type 2 diabetes prescribing patterns changed during the pandemic and whether these changes differed across socioeconomically deprived areas.

#####Note: COVID period is classed from March 2020 - June 2021, theriod between when lockdown was introduced, and when restrictions were fully lifted

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor) # cleaning data
library(gt) # tables
library(here) # directory structure
library(readxl) # read in excel file
library(sf) #to read in map data
library(plotly) #to display information when plot is hovered over

#Get data from 24-month period from Mar 2020 - Feb 2022 
#create function to read data in:
prescriptions_data <- function() {
  files <- c(paste0(c("jan", "feb","mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"), "2020"),
             paste0(c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"), "2021"))
  
  map_dfr(files, ~read_csv(here("data", paste0(.x, ".csv"))) %>%
                   clean_names() %>%
                   mutate(file_name = .x))}

# Read in data
data_for_years <- prescriptions_data()
deprivation_stats <- read_excel(here("data/SIMD.xlsx"))
```

```{r message=FALSE, warning=FALSE}
#Read in and tidy healthboard population database
population_data <- read_csv(here("data/general_health_census.csv"), skip = 10) %>% 
  filter(!is.na('All people')) %>%  #Removes NA
  rename(hb_name = "General health",
         hb_population = "All people") %>% 
  select(hb_name, hb_population) %>% 
  filter(!str_detect(hb_population, "Cells")) #Removes rows with text in population (to remove copyright info)
```

```{r message=FALSE, warning=FALSE}
diabetes_summary <- data_for_years %>%
  #filter for diabetes medications
  filter(!is.na(bnf_item_description)) %>% 
  filter(str_detect(bnf_item_description,
                    regex("metformin|gliptin|gliflozin|glutide|glitazone|gliclazide|glimepiride", ignore_case = TRUE))) %>% 
    group_by(paid_date_month, hbt) %>% 
    summarise(paid_quantity = sum(paid_quantity, na.rm = TRUE)) %>% 
#change months from numbers to letters
    mutate(paid_date_month = ym(paid_date_month))

# Merge all tables together
combined_health_data <- deprivation_stats %>%
  full_join(diabetes_summary, join_by(HBcode == hbt), relationship = "many-to-many") %>%
  left_join(population_data, join_by(HBname == hb_name))

# Now calculate prescriptions per person
combined_health_data <- combined_health_data %>%
  group_by(HBname, paid_date_month) %>%
  mutate(quantity_per_head = sum(paid_quantity, na.rm = TRUE) / mean(as.numeric(hb_population))) #calculate prescriptions per person and change hb_population to numeric value instead of character
```

```{r message=FALSE, warning=FALSE}
# Creating a summary table and filtering for common type 2 medications
totalmeds_plot <- combined_health_data %>% 
  ggplot(aes(x = paid_date_month, y = quantity_per_head)) +
  # Add vertical lines for COVID start and end
  geom_vline(xintercept = as.numeric(ymd("2020-03-01")), 
             linetype = "dashed", colour = "purple", size = 0.8) +
  geom_vline(xintercept = as.numeric(ymd("2021-06-01")), 
             linetype = "dashed", colour = "purple", size = 0.8) +
  geom_line(colour = "darkblue", size = 0.5) + # Make original line lighter
  labs(
    title = "Prescription of type 2 dabetes medications per person",
    subtitle = "January 2020 - December 2021 | Purple dashed lines: COVID period (Mar 2020 - Jun 2021)",
    x = "Month",
    y = "Quantity per head",
  ) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(~HBname, scales = "free_y")
#labeller = label_wrap_gen(width = 10) ADD TO FACET WRAP

print(totalmeds_plot)

#MENTION FREED SCALES as there were vast differences in prescription across healthboards
```

##### An overall upward trend was seen during COVID and this increase accelerated post COVID. Changes in the trend during the COVID perios may be due to the enforcement and lifting of lockdown rules. To further examine and link trends to deprived areas. I will be using January 2020 (baseline), January 2021 (third lockdown), December 2022 (post pandemic/current trend). - I chose these seasons to balance temporal evidence (pandemic stages) and seasonal consistency (comparing winter months avoids confounding by seasonality)


```{r message=FALSE, warning=FALSE}
key_months_data <- combined_health_data %>% 
  filter(paid_date_month %in% as.Date(c("2020-01-01", "2021-12-01"))) %>% 
  group_by(HBname) %>% 
  mutate(covid_phase = case_when(
    paid_date_month == as.Date("2020-01-01") ~ "pre_covid",
    paid_date_month == as.Date("2021-12-01") ~ "post_covid"))

# Summarise by health board and COVID phase
lollipop_data <- key_months_data %>%
  group_by(HBname, covid_phase) %>% 
  summarise(avg_qph = mean(quantity_per_head, na.rm = TRUE),
            avg_simd = mean(SIMD2020v2_Decile, na.rm = TRUE)) %>%
  # Pivot to get Pre and Post columns for connecting lines
  pivot_wider(names_from = covid_phase, values_from = avg_qph) %>%
  mutate(percent_change = ((post_covid - pre_covid)/pre_covid)*100) %>% #to calculate percent change
  # Arrange by descending avg_simd
  arrange(desc(avg_simd)) %>%
  ungroup() %>% #before creating a factor so hat HBname can be arranged in order of avg_simd in graph
  # Create factor to preserve ordering in plot
  mutate(HBname = factor(HBname, levels = HBname))

lollipop_graph <- lollipop_data %>% 
  ggplot() +
  geom_segment(aes(x = HBname, xend = HBname, y = pre_covid, yend = post_covid), 
               color = "grey") +
  geom_point(aes(x = HBname, y = pre_covid), 
             color = rgb(0.2, 0.7, 0.1, 0.5), size = 3) +
  geom_point(aes(x = HBname, y = post_covid), 
             color = rgb(0.7, 0.2, 0.1, 0.5), size = 3) +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none") +
  xlab("") +
  ylab("Average Quantity Per Head") +
  ggtitle("Change in Quantity Per Head: Pre-COVID vs Post-COVID",
          subtitle = "NHS Health Boards ordered by average SIMD decile (descending)")

# Display the plot
print(lollipop_graph)
```


```{r message=FALSE, warning=FALSE}
# load the NHS Health board Shapefile
NHS_healthboards <- st_read(here("data/Week6_NHS_HealthBoards_2019.shp"))

#Join spatial data with other health data
summary_map <- NHS_healthboards %>%
  full_join(lollipop_data, by = join_by(HBName == HBname))
```

```{r message=FALSE, warning=FALSE}
# Plot percentage change map
plot_map <- summary_map %>% 
  ggplot() +
  geom_sf(aes(fill = percent_change, text = paste0("Health Board: ", HBName, "\n", round(percent_change,2), "%")), color = "white", size = 0.3) +
  scale_fill_viridis_c(option = "plasma", 
                       name = "% Change in Average Quantity per Head") +
  labs(
    title = "Percentage Change in Average Type 2 Diabetes Prescriptions Per Head \nPre-COVID → Post-COVID",
    caption = "Data from Jan 2020 and Dec 2021"
  ) +
  theme_minimal() +
  theme(legend.position = "right", plot.title = element_text(face = "bold"))

interactive_map <- ggplotly(plot_map, tooltip = "text")

interactive_map
```


```{r message=FALSE, warning=FALSE}
ranked_table <- lollipop_data %>%  # replace with your dataframe name
  arrange(avg_simd) %>% 
  mutate(Rank = row_number()) %>% 
  select(c(Rank, HBname, pre_covid, percent_change))

# ---- Step 2: GT table ----
ranked_table %>%
  gt() %>%
  cols_label(
    Rank = "Rank",
    HBname = "Health Board",
    pre_covid = "Pre-COVID Level (per head)",
    #absolute_change = "Absolute Increase",
    percent_change = "Percentage Increase (%)"
  ) %>%
  fmt_number(
    columns = c(pre_covid),
    decimals = 2
  ) %>%
  fmt_number(
    columns = percent_change,
    decimals = 1
  ) %>%
  tab_header(
    title = " Health Boards by Percentage Increase in Type 2 Diabetes Prescriptions",
    subtitle = "Comparing pre-COVID (Jan 2020) to post-COVID (Dec 2021)"
  )
```

```

##MAKE DATA PER 100,000 PEOPLE ++ Add what line says, + DOUBLE CHECK LABELS + HOVER OVER MAP SHOULD SAY HEALTHBOARD NAME

#### Questions:

#### 1.  In feedback what do you mean by showing links for reproducibility? Is code enough? Or do we need to show/write how we got to the links? #### Write overall urls
#### 2.  For deprived areas, is an overall view of healthboards enough or would you want to see simd inside datazones? classify each healthboard as a colour, average prescriping would get lighter or darker
#### 3.  How do I make the figures wider so it doesn't cut off names?
